---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fdakmapp

<!-- badges: start -->
[![R build status](https://github.com/astamm/fdakmapp/workflows/R-CMD-check/badge.svg)](https://github.com/astamm/fdakmapp/actions)
[![Coverage](https://github.com/astamm/fdakmapp/workflows/test-coverage/badge.svg)](https://github.com/astamm/fdakmapp/actions)
[![Website](https://github.com/astamm/fdakmapp/workflows/pkgdown/badge.svg)](https://github.com/astamm/fdakmapp/actions)
[![Render README](https://github.com/astamm/fdakmapp/workflows/render-readme/badge.svg)](https://github.com/astamm/fdakmapp/actions)
<!-- badges: end -->

The **fdakmapp** package provides the `kma()` function that jointly performs clustering and alignment of a functional data set. Mean and medoid center methods are available. The algorithm is fully parallelized.

## Getting Started

The source code can be cloned or downloaded directly from github.
An R studio project file is provided to open the project in RStudio.

### Prerequisites

The package is linked against OpenMP, the BLAS and LAPACK libraries. Therefore, you should make sure that your R / RStudio environment is ready

- to build packages: to check this, type in the console
```
# install.packages("pkgbuild")
pkgbuild::has_devel()
```

- to build packages with compiled code: 

    - on Linux: it should be ready;
    - on Windows: you should install [Rtools](https://cloud.r-project.org/bin/windows/Rtools/);
    - on macOS: you should install a [set of tools](https://cloud.r-project.org/bin/macosx/tools/).

- to build packages with C++ code through Rcpp: on macOS, you can for instance follow this [tutorial](https://thecoatlessprofessor.com/programming/cpp/r-compiler-tools-for-rcpp-on-macos/).

### Installation

The **fdakmapp** package is currently only available on GitHub. You can install it as follows:
```
# install.packages("remotes")
remotes::install_github("astamm/fdakmapp")
```

### Example

```{r example-simulated30}
library(fdakmapp)

res <- kma(
  simulated30$x,
  simulated30$y,
  seeds = c(1, 11, 21),
  n_clust = 3,
  center_method = "medoid",
  warping_method = "affine",
  dissimilarity_method = "pearson"
)

plot(res)
```
