---
title: "Supported input formats"
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
load("../../R/sysdata.rda")
```

```{r setup}
library(fdacluster)
library(ggplot2)
```

We shall recall that
[**fdacluster**](https://astamm.github.io/fdacluster/index.html) handles only
functional data on univariate domains. This is because alignment (a.k.a.
registration) of curves defined on multivariate domains is much more complex.
The codomain of the functional data can however very well be multivariate.
Throughout this article, we will consistently denote:

- by $N$ the number of curves in the sample, which is the sample size;
- by $M$ the size of the grids on which the curves have been evaluated;
- by $L$ the dimension of the codomain of the curves.

## Native format

The native format in
[**fdacluster**](https://astamm.github.io/fdacluster/index.html) consists in a
pair of vector/matrix `x` and matrix/array `y` respectively storing the
evaluation grids and the values of the curves on these grids.

### Passing evaluation grids through the `x` argument

If all curves have been evaluated on the exact same grid, then `x` can be either
a numeric vector of size $M$ or a numeric matrix of size $N \times M$, in which
case all rows should be identical.

It is however possible that, despite having a *common grid size*, curves have
actually been evaluated on different grids. In this case, `x` should be a matrix
of size $N \times M$.

### Passing curve values through the `y` argument

If the curves evaluate on a univariate codomain, then `y` can be either a
numeric matrix of size $N \times M$ or a numeric array of size $N \times 1
\times M$. In both cases, it stores individual curves in rows and their value on
the corresponding grid point in either the second dimension if `y` is a matrix
or the third dimension if `y` is an array.

It is however possible that curves evaluate on a multivariate codomain of
dimension $L$ In this case, `y` should be an array of size $N \times L \times
M$.

## Parsers for standard functional R objects

The reason for such a choice of native format is computation efficiency.
However, the [R landscape of functional data
analysis](https://cran.r-project.org/view=FunctionalData) is very
rich. This has led us to make
[**fdacluster**](https://astamm.github.io/fdacluster/index.html) compatible with
other standard ways of storing functional data in R. Specifically, we added
support for the `fd` class from the
[**fda**](https://cran.r-project.org/package=fda) package and for the `funData`
and `multiFunData` classes from the
[**funData**](https://cran.r-project.org/package=funData) package.

### Support for [**funData**](https://cran.r-project.org/package=funData)

The [**funData**](https://cran.r-project.org/package=funData) package is a
package designed for the sole purpose to implement classes for storing all kinds
of functional data. Functional data have a domain and a codomain, both of which
can be univariate or multivariate. In addition, each function in the data sample
might have been observed on a different grid. This has led the authors of the
[**funData**](https://cran.r-project.org/package=funData) package to propose
three different classes, namely [@happ2020object]:

> 
1. Class `funData` for dense functional data of *arbitrary* dimension (in most
cases the dimension of the domain is $d \in \{1, 2, 3\}$) on a common set of
observation points $t_1, \dots, t_S$ for all curves. The curves may have missing
values coded by `NA`.
2. Class `irregFunData` for irregularly sampled functional data with individual
sampling points $t_{ij}$, $j = 1, \dots, S_i$, $i = 1, \dots ,N$ for all curves.
The number $S_i$ and the location of observation points can vary across
individual observations. At the moment, only data on one-dimensional domains is
implemented.
3. Class `multiFunData` for multivariate functional data, which combines $p$
elements of functional data that may be defined on different dimensional domains
(e.g., functions and images).

The [**fdacluster**](https://astamm.github.io/fdacluster/index.html) package
focuses on functional data defined on uni-dimensional domains with possibly
multivariate codomains and possibly different evaluation grids per curve. As
such, support for all three classes is needed. This is achieved by the utility
function `format_inputs()` which can read in classes from the
[**funData**](https://cran.r-project.org/package=funData) package when provided
as `x` argument with no `y` argument. Specifically, the following code shows how
classes from the [**funData**](https://cran.r-project.org/package=funData)
package are transformed into the `(x, y)` matrix / array format:

```r
if (inherits(x, "funData")) {
  if (length(x@argvals) > 1)
    cli::cli_abort(c(
      "The {.pkg fdacluster} package does not support functional data defined ",
      "on multivariate domains."
    ))
  L <- 1
  y <- x@X
  dims <- dim(y)
  N <- dims[1]
  M <- dims[2]
  y <- array(y, dim = c(N, L, M))
  x <- x@argvals[[1]]
} else if (inherits(x, "irregFunData")) {
  L <- 1
  N <- length(x@argvals)
  M <- x@argvals |> 
    sapply(length) |> 
    mean() |> 
    round()
  y <- array(dim = c(N, L, M))
  y[, 1, ] <- x@X |> 
    imap(\(values, id) approx(x@argvals[[id]], values, n = M)$y) |> 
    do.call(rbind, args = _)
  x <- x@argvals |> 
    lapply(\(grid) seq(min(grid), max(grid), length.out = M)) |> 
    do.call(rbind, args = _)
} else if (inherits(x, "multiFunData")) {
  L <- length(x)
  dims <- dim(x[[1]]@X)
  grid <- x[[1]]@argvals[[1]]
  lapply(x, \(fData) {
    if (length(fData@argvals) != 1)
      cli::cli_abort(c(
        "The {.pkg fdacluster} package does not support functional data ", 
        "defined on multivariate domains."
      ))
    if (any(fData@argvals[[1]] != grids))
      cli::cli_abort(c(
        "All components of the {.cls multiFunData} object must share the same ",
        "evaluation grids."
      ))
    if (any(dim(fData@X) != dims))
      cli::cli_abort(c(
        "All components of the {.cls multiFunData} object must have values ", 
        "stored in matrices with the same dimensions."
      ))
  })
  N <- dims[1]
  M <- dims[2]
  y <- array(dim = c(N, L, M))
  for (l in 1:L) y[, l, ] <- x[[l]]@X
  x <- grid
} else
  cli::cli_abort(c(
    "Functional data provided in a single argument {.arg x} must be either of ",
    "class {.cls funData} or of class {.cls irregFunData} or of class ", 
    "{.cls multiFunData}."
  ))
```

#### Example for `funData` support: Growth Data

The following lines of code demonstrates how clustering functions in
[**fdacluster**](https://astamm.github.io/fdacluster/index.html) handle objects
of class `funData` using the Berkeley growth data study developed in
@silverman2002applied.

```{r, eval=FALSE}
growth <- fda::growth
growthData <- funData::funData(
  argvals = growth$age, 
  X = t(cbind(growth$hgtm, growth$hgtf))
)
out_growth <- fdakmeans(
  x = growthData,
  n_clusters = 2,
  seeding_strategy = "exhaustive-kmeans++",
  cluster_on_phase = TRUE,
  use_verbose = FALSE
)
```

```{r, fig.width=7, fig.height=5, fig.retina=2}
plot(out_growth)
```

#### Example for `irregFunData` support: Simulated Data

The following lines of code demonstrates how clustering functions in
[**fdacluster**](https://astamm.github.io/fdacluster/index.html) handle objects
of class `irregFunData` using in-house simulated data on the fly.

```{r, eval=FALSE}
# Generates full grid
argvals <- seq(0, 2 * pi, by = 0.01)
# Simulate 30 irregular grids with various sampling points (number and values)
indices <- replicate(30, sort(sample(1:length(argvals), sample(30:50, 1))))
argvalsIrreg <- lapply(indices, \(i) argvals[i])
# Simulate functional data with obvious grouping structure
withr::with_seed(1234, {
  simData <- funData::irregFunData(
    argvals = argvalsIrreg, 
    X = mapply(
      function(x, a, b) a * sin(x + b), 
      x = argvalsIrreg, 
      a = c(rgamma(10, 25, 50), rgamma(10, 50, 50), rgamma(10, 100, 50)),
      b = c(rnorm(10, -1, 0.1), rnorm(10, 0, 0.1), rnorm(10, 1, 0.1))
    )
  )  
})
out_sim <- fdakmeans(
  x = simData,
  n_clusters = 3,
  seeding_strategy = "exhaustive-kmeans++",
  use_verbose = FALSE
)
```

```{r, fig.width=7, fig.height=5, fig.retina=2}
plot(out_sim)
```

#### Example for `multiFunData` support: Gait Data

The following lines of code demonstrates how clustering functions in
[**fdacluster**](https://astamm.github.io/fdacluster/index.html) handle objects
of class `multiFunData` using the gait data study developed in
@silverman2002applied.

```{r, eval=FALSE}
cycle_perc <- (0:19) / 19 * 100
hipData <- t(fda::gait[, , 1])
hipData <- hipData
kneeData <- t(fda::gait[, , 2])
kneeData <- kneeData
gaitData <- funData::multiFunData(
  funData::funData(argvals = cycle_perc, X = hipData), 
  funData::funData(argvals = cycle_perc, X = kneeData)
)
out_gait <- fdakmeans(
  x = gaitData,
  n_clusters = 1,
  seeding_strategy = "exhaustive",
  warping_class = "srsf",
  use_verbose = FALSE
)
```

```{r, fig.width=7, fig.height=5, fig.retina=2}
p <- autoplot(out_gait)
p$data$component_id <- factor(p$data$component_id, labels = c("Hip", "Knee"))
p + labs(y = "Angle (degree)", title = "Gait Kinematics")
```

### Support for [**fda**](https://cran.r-project.org/package=fda)

The [**fda**](https://cran.r-project.org/package=fda) package features the class
`fd` as a hosting `R` structure for functional data. In this representation,
functional data are decomposed on a number of basis functions
[@ramsay2006functional]. The class is then a list with the following 3
components:

- `coefs`: Stores the coefficients of each functional datum in the sample
corresponding to its expansion along the basis functions;
- `basis`: Stores the basis functions;
- `fdnames`: Stores IDs for each coefficient, each observation and each 
dimension of the co-domain.

Basis functions supported by the
[**fda**](https://cran.r-project.org/package=fda) package are exclusively
defined on univariate domains. As such, the class `fd` only handles functional
data on unidimensional domains. However, it accommodates functional data
evaluating on multidimensional co-domain. This is achieved by using a
three-dimensional array for the coefficients instead of a vector or a matrix.
Indeed, quoting the help page of the `fd` class:

>
`coef`: A vector, matrix, or three-dimensional array of coefficients. The first
dimension (or elements of a vector) corresponds to basis functions. A second
dimension corresponds to the number of functional observations, curves or
replicates. If `coef` is a vector, it represents only a single functional
observation. If `coef` is an array, the third dimension corresponds to variables
for multivariate functional data objects. A functional data object is
*univariate* if `coef` is a vector or matrix and *multivariate* if it is a
three-dimensional array.

Consequently, the
[**fdacluster**](https://astamm.github.io/fdacluster/index.html) package handles
objects of class `fd` by considering that `x`, as in native format, stores the
unidimensional grids and `y` is an object of class `fd`. It then parses the data
in the following way:

```r
dims <- sapply(y$fdnames, length)
M <- dims[1]
N <- dims[2]
L <- dims[3]
if (is.vector(x)) {
  if (length(x) != M)
    cli::cli_abort(c(
      "The number of function evaluations ({M}) does not match the grid ",
      "size ({length(x)})."
    ))
  y <- fda::eval.fd(x, y)
} else {
  if (nrow(x) != N)
    cli::cli_abort(c(
      "When provided multiple evaluation grids as a matrix, the number of ",
      "rows should match the number of curves."
    ))
  if (ncol(x) != M)
    cli::cli_abort(c(
      "When provided multiple evaluation grids as a matrix, the number of ",
      "columns should match the common grid size."
    ))
  y <- fda::eval.fd(t(x), y)
}
if (is.null(dim(y))) {
  # y is a single 1-dimensional curve
  y <- array(y, dim = c(length(y), 1, 1))
} else if (length(dim(y)) == 2) {
  # y is N 1-dimensional curves
  y <- array(y, dim = c(dim(y), 1))
}
y <- aperm(y, c(2, 3, 1))
```

#### Example for `fd` support: Gait Data

The following lines of code demonstrates how clustering functions in
[**fdacluster**](https://astamm.github.io/fdacluster/index.html) handle objects
of class `fd` using the gait data study developed in @silverman2002applied.

```{r, eval=FALSE}
bspl <- fda::create.fourier.basis(rangeval = c(0, 100), nbasis = 5)
gaitDataFD <- fda::smooth.basis(
  argvals = cycle_perc, 
  y = fda::gait, 
  fdParobj = bspl
)$fd
out_gait_fd <- fdakmeans(
  x = cycle_perc,
  y = gaitDataFD,
  n_clusters = 1, 
  seeding_strategy = "exhaustive",
  warping_class = "srsf",
  use_verbose = FALSE
)
```

```{r, fig.width=7, fig.height=5, fig.retina=2}
p <- autoplot(out_gait_fd)
p$data$component_id <- factor(p$data$component_id, labels = c("Hip", "Knee"))
p + labs(y = "Angle (degree)", title = "Gait Kinematics")
```

## References
